Gradual Soundness: Lessons from Static Python

### 2022-04-29 pj-7.1/response.txt

Fixed `str` and removed `object`.
[p.12 s.4.1]

(`object` was there as a simple instance of the `Object` type. Since it's not
needed to explain type boundary issues, we can remove it.)

> Line 6: `C.f(expr)`
> * looks like a static method call to me, is that right? That's usually
>   trivial compared with the dynamic dispatch of a real method call.
> * how do you invoke a method on an object?

Oh dear. This was not meant to be a static method call. We were far too
aggressive when simplifying from the Redex model.

The revision uses `expr.f(expr)` instead and looks up the class type of the
receiver to type check the call.
[p.12 s.4.1]

@@

We have reworded the caption.
[p.15 s.4.2.1]

> ### p14, fig 5
>
> * rule `M-App` seems too limiting. Literally read, it's not possible to
>   invoke inherited methods because the method `f0` must be defined in class
>   `C0` in which it is defined. This rule affirms the class-method conjecture
>   from above.

We had envisioned the lookup `Env(C0)` as searching both `C0` and its parents
for all visible fields and methods.

The revised figure writes `Env*(C0)` instead and includes a note to this effect.
(The revision also uses `\Gamma` instead of `Env`)
[p.14 s.4.2.2]

@@

Thanks!
[p.14 s.4.2.2]

> ### p15, -16
>
> *better left to the Redex mechanization*
>
> It would be useful to show the excerpt for method calls here, as they seem to
> be nonstandard.

Method calls in Redex are standard; they work like calls in (Static) Python.
The confusion here is due to our over-simplified model in the submission, which
gave the impression that only a class can receive a method call.

The revised syntax for the model addresses this.
[p.12 s.4.1]

@@

> Page 13 line -7: A dynamically-typed variable -> A dynamically-typed function

No, "variable" is there on purpose because we don't know statically what a
dynamically typed variable points to. It might not be a function.

> Figure 5 rule D-App: in Figure 3 a function f seems not to be allowed to have
> type Dyn in the enviroment

This is allowed, but only indirectly. The definition of a function `f_0` cannot
directly say that `f_0` has type Dyn, but it can declare the type `Dyn -> Dyn`
and then apply `f_0(f_0)` to assign the dynamic type to a function.

@@

(Note that `x` and `f` are undifferentiated variable names. The model uses `x`
 in some definitions and `f` in others as hints to readers, but it's fine to
 name a function `x_2` and so on.)

> Also, Object seems to be the top type, hence also Bool and Int are seen as
> classes, right?

Yes, Bool and Int are technically classes. The paper does not discuss this,
but the Redex model (and Static Python) give them full support as classes.

### 2022-05-16 pj-7.1/paper.tex
pj: add section on gradual soundness

%% TODO for camera-ready
%% -

@@

\SP{} is part of a large codebase that includes a type system, a tailored
bytecode, and a method-based JIT compiler.
In essence, \SP{} is the type system for an entirely new language.
The interface that it offers to programmers, however, replicates the
standard Python experience.
\SP{} runs Python 3.8 programs with minimal changes to their
behavior~(\cref{s:not-python}),
and it compiles code on-the-fly to support existing
IDEs and developer tools.
The advanced features of the \SP{} type system are offered on an opt-in basis
(\emph{gradual soundness})
and arranged so that programmers can begin adding types one module at a time.

@@

>>>>

%% TODO for camera-ready
%% - reviews-r2: clarify gradual soundness again, what do we mean?
%%    (1) two layers of types+code for some algorithms ... esp. for data structures ... key to practical concrete
%%    (2) progressive types for other code
%%    ... in short, progressive types adapted to GT
%%   - try section 2 for this

@@

\SP{} is part of a large codebase that includes a type system, a tailored
bytecode, and a method-based JIT compiler.
In essence, \SP{} is the type system for an entirely new language.
The interface that it offers to programmers, however, replicates the
standard Python experience.
\SP{} runs Python 3.8 programs with minimal changes to their
behavior~(\cref{s:not-python}),
and it compiles code on-the-fly to support existing
IDEs and developer tools.
The advanced features of the \SP{} type system are offered on an opt-in basis
(\emph{gradual soundness},~\cref{s:gradual-soundness})
and arranged so that programmers can begin adding types one module at a time.

@@

\subsection{Gradual Soundness}
\label{s:gradual-soundness}

\SP{} represents a new approach to the design of gradually-typed
languages.
Whereas prior works choose to either force changes upon untyped code
with \emph{concrete} types~\cite{wnlov-popl-2010,mt-oopsla-2017}
or accommodate untyped code as-is with \emph{migratory}
types~\cite{st-sfp-2006,tfffgksst-snapl-2017}, \SP{} provides a multi-level
type system that supports both styles without expensive run-time checks.
Shallow types accommodate Python code, but offer little in the way
of type-based reasoning and type-directed optimization.
Concrete types and primitive types give stronger type guarantees, but have
drawbacks:
concrete types force changes upon untyped code and primitive types
add restrictions to typed code.
The overall system, in which programmers can choose between untyped code
and several increasingly-restrictive typed languages,
is a combination of gradual typing and progressive
types~\cite{pqk-onward-2012} that we call \emph{gradual soundness}.

To be clear, gradual soundness has a different focus than gradual typing.
Gradual typing in the traditional, broad sense is about finding a best-of-both worlds
combination of typed and untyped
code~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006}.
Gradual soundness fits under this broad umbrella, and contributes the idea that multiple
type systems may be needed for an optimal mix.
Gradual typing in the refined sense calls for a dynamic type
that satisfies the gradual guarantees~\cite{svcb-snapl-2015}.
A gradually-sound language may choose to satisfy these guarantees.
\SP{}, however, does not~(\cref{s:type-dynamic}).
